Determining Causes of Death with ChatGPT: A Case Study of Deaths from Verbal Autopsy in Sierra Leone from 2019-2022

### 2024-01-03 manuscript/manuscript.tex
Working on background

Reliable counts of deaths and diagnosis of their causes are crucial for public health planning and policy making --- guiding interventions and resource allocation to reduce premature deaths \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries have reported to have either no data on deaths or have registered less than 50\% of deaths in their country, where an even fewer 8\% of these registered deaths have a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill in this gap, an alternative method known as Verbal Autopsy (VA) is used to determine likely causes for deaths at scale and outside of traditional healthcare facilities \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}. VA involves trained surveyors that interview those familiar with the deceased to gather information using standardized questionnaires and open narratives, while medical professionals evaluate the gathered information to assign probable CODs \cite{worldhealthorganizationVerbalAutopsyStandards2023}.

>>>>

Reliable counts and diagnosis of deaths are crucial for public health planning and policy making --- guiding scalable interventions that save lives and reduce premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries have reported to have either no data on deaths or have registered less than 50\% of deaths in their country, where an even fewer 8\% of these registered deaths have a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill in this gap, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale, outside of traditional healthcare facilities \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}.

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, family, relatives, friends) to gather information using standardized questionnaires and open narratives. The gathered information consists of a standardized questionnaire, and an open narrative. In the COD assignment component, physicians evaluate information gathered from lay surveyors to assign probable CODs. Although the survey component has been an effective alternative to collect mortality data at scale, the COD assignment component has been criticized to be expensive and difficult to reproduce due to reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. Recently, computer algorithms have been studied to automatically assign CODs as an alternative to physicians, with performances close to physicians at the population level, but poor performances at the individual level ((CITE)). Many of these computer algorithms have utilized the questionnaire portion of the surveyed data, but often omit data from the open narrative.
